---
title: "PS6"
output: html_document
date: "2023-11-27"
author: "Matthew Zhuo"
---
### 1

#### 1.1

##### 1.1.1
```{r}
library(magrittr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
data <- read.delim("scripps-co2.csv.bz2")
data <- data %>%
  select(year,month,co2)
head(data)
```

##### 1.1.2
Each line represents the concentration of co2 in ppm for each month in a time period

##### 1.1.3
Missing values are recorded as -99.99 in this dataset
```{r}
for (i in 1:length(data$co2)) {
  if (data$co2[i] < 0) {
    print(data$co2[i])
  }
}
```


##### 1.1.4
```{r}
data$year %>%
  min()
data$year %>%
  max()
```


##### 1.1.5
```{r}
data <- data[data$co2 != -99.99,]
data$date <- make_date(year=data$year,month=data$month)
data %>%
  ggplot(aes(date,co2)) + geom_point(size=0.5) + geom_line() + labs(title="Co2 concentration from 1958 to 2023",subtitle = "From the scripps-co2 dataset",x="year",y="Co2 concentration in ppm")

```

##### 1.1.6
The co2 concentration has increased over time

#### 1.2

###### 1.2.1
```{r}
Tempdata <- read.delim("hadcrut-5.0.1.0-annual.csv.bz2")
head(Tempdata)
```

##### 1.2.2
A line represents the degree in celsius and the confidence levels for that year

##### 1.2.3
```{r}
Tempdata$Time %>%
  min()
Tempdata$Time %>%
  max()
```

##### 1.2.4
```{r}
preindustrial <- Tempdata %>%
  filter(Time<=1900) %>%
  summarize(preindustrial_avg=mean(Anomaly..deg.C.))
preindustrial
secondAvg <- Tempdata %>%
  filter(Time>=1961&Time<=1990) %>%
  summarize("1961 to 1990 avg"=mean(Anomaly..deg.C.))
secondAvg
Tempdata$anomaly_diff <- Tempdata$Anomaly..deg.C.-secondAvg$`1961 to 1990 avg`
head(Tempdata)
```

##### 1.2.5
```{r}
library(ggplot2)
Tempdata %>%
  ggplot(aes(Time,anomaly_diff)) + geom_point() + labs(title="how much Co2 concentration increased from 1850-2023",subtitle="From the hadcrut dataset",x="year",y="increase in Co2 concentration")
```

###### 1.2.6
Co2 has increased over time

#### 1.3

##### 1.3.1
Time/year

##### 1.3.2
you combine the data from each month into a year then combine the datasets

##### 1.3.3
```{r}
data <- data %>%
  group_by(year) %>%
  summarize(avg_co2=mean(co2))
Tempdata <- Tempdata %>%
  rename(year=Time)
mergedData <- right_join(data,Tempdata,by="year")
head(mergedData)
```

##### 1.3.4
```{r}
mergedData %>%
  ggplot(aes(avg_co2,anomaly_diff)) + geom_point(aes(color=year)) + geom_smooth(method=lm,se=FALSE) + geom_hline(yintercept = 1.5,color="red") + labs(title="Co2 concentration vs Temperature",subtitle="From the hadcrut dataset",x="Co2 concentration ppm",y="Temperature anomaly (w.r.t 1850-1900 average) Celcius",color="decade")
```

### 2

##### 2.1
```{r}
uahData <- read.delim("UAH-lower-troposphere-wide.csv.bz2")
uahData <- uahData %>%
  select(year,month,globe)
uahData %>%
  head()
```

##### 2.2
```{r}
uahDataAvgs <- uahData %>%
  group_by(year) %>%
  summarize(uahAvg=mean(globe))
head(uahDataAvgs)
```

##### 2.3
```{r}
Mergeddata <- Tempdata %>%
  right_join(uahDataAvgs)
Mergeddata <- Mergeddata %>%
  select(year,anomaly_diff,uahAvg)
Mergeddata <- Mergeddata %>%
  rename("lower trophsphere"=uahAvg,surface=anomaly_diff)
head(Mergeddata)
```

##### 2.4
wide form

##### 2.5
```{r}
Mergeddata <- Mergeddata %>%
  pivot_longer(c(surface,`lower trophsphere`),names_to = "type",values_to = "anomaly")
```


##### 2.6
```{r}
Tempdata <- Tempdata %>%
  filter(year >= 1980)
Tempavg <- Tempdata %>%
  summarize(avg=mean(Tempdata$Anomaly..deg.C.))
Mergeddata$Temp_anomaly <- Tempdata$Anomaly..deg.C.-Tempavg
Mergeddata %>%
  ggplot(aes(year,anomaly)) + geom_point(aes(col=type)) + geom_smooth(data=subset(Mergeddata,Mergeddata$type != "surface"),method=lm,se=FALSE,color="red") + geom_smooth(data=subset(Mergeddata,Mergeddata$type != "lower trophsphere"),method=lm,se=FALSE,color="lightblue") + labs(title="temperature anomaly changes", subtitle="From the hadcrut and UAH datasets", x="year",y="temperature anomaly")
```

##### 2.6.a
The surface has a higher increase in temperature over this time period